\chapter*{Abstract}


We study transport and spreading in sparse disordered networks. The networks are 
constructed by assigning spatial locations to sites. Distance matrices (also called
Euclidean Distance Matrices) form the basis to define either rate equations or Hamiltonians.
Additionally, the matrix might depend on some coupling parameter assigned to each network edge.
Thus, by randomizing either the locations of the sites
or the coupling parameter, we can induce disorder.



By allowing the network elements to differ by orders of magnitude,
with most of the elements vanishingly small, we create a "sparse" network.
Our interest lies in the long term transport properties of these sparse systems. 


We approach the subject of transport by looking at the spectral properties
of these networks, as well as by their resistor network analogies. For stochastic systems we see 
that in pure $1d$ this sparsity causes subdiffusion, while in quasi $1d$ and
higher dimensions the system remains diffusive but with a reduced diffusion coefficient. 


The \emph{ERH - Effective Range Hopping -} method is suggested to estimate this diffusion 
coefficient by combining ideas from the well established \emph{VRH} scheme with 
ideas from percolation theory.


The numerical results for the diffusion coefficient are compared to the linear
estimate and to the improved \emph{ERH} result.


Further generalizations are discussed, including multiple interacting particles and
an assymetric version to the VRH procedure. Another proposed model is a $2d$ variant of 
the Sinai model.
