\chapter{Objectives}

%TODO: orientation.

The first two objectives have been the target of my MSc, and are covered
in an article published at Physical Review E \cite{de_leeuw_diffusion_2012},
and reprinted here in \autoref{sec:papers}. 
The other objectives will be pursued during my PhD research.


%%%%%%%%%
\section{Diffusion or Subdiffusion in $2d$}

For the stochastic random network, the question of diffusion in most $1d$
systems was analytically solved in \cite{alexander_excitation_1981}. 
The $2d$ case is, as far as we know, not yet analytically solved, and is much less clear. 
In \cite{amir_localization_2010} it is claimed that in low density systems subdiffusion of
order $\sim log^d$ should occur. We wish to examine these networks through their spectral
properties to find out if this is indeed the case. Our focus will be on low densities,
and on checking whether there exists a transition between low density and high density behavior.

Another aspect we wish to investigate is the influence of geometrical and 
topological features on the physics of the system, beyond the direct 
effect of changing the matrix elements statistics.

%%%%%%%%%
\section{Banded sparse matrices}


Between the $1d$ nearest neighbor network and the $2d$ network,
lie the quasi $1d$ network, which allows transitions to further neighbors.
The conductance of quasi-$1d$ banded sparse matrices was studied
numerically in \cite{stotland_random-matrix_2010}.
There, they generalized the \emph{Variable-Range-Hopping} scheme to treat 
matrices in the sparse regime, where 
$(\text{sparsity}\cdot \text{bandwidth}) \ll 1$. 
However, in this work it is not clear what are the limits on either
sparsity or bandwidth, and in particular where is the cross between the 
validity regime of VRH and that of SLRT. We will try to understand this
issue analytically.



% TODO A specific version of these matrices were also studied in \cite{janssen_correlated_2000}.

%%%%%%%%%%%
\section{Models with relaxation}

%% same potentials - is not relevant.
%% do I mean Steady State?
% asymmetric transitions
% Boltzman - detailed balance
%% remove asymmetric VRH
% models with relaxations

% the point is that I want to deal with asymmetry.
%% "generaly the transition matrices are asymmetric,
%% but up to now we dealt with symmetric only."
%% why combine on-site energy with asymmetry.

% Three categories:
% wnm = wmn  - no talk about steady state - equilibrium transient diffusion
%%            resnet
% wnm propto wmn - detailed balance - stochastic field
%           integral over it must be zero  boltzmann equil
%           should be added to the background
% wnm \ne wmn - asymmetry - frustrated NESS

Up to now we have dealt with symmetric matrices only, but as presented in 
the introduction \autoref{sec:matrices}, in general the transition 
rates may be asymmetric. We wish to extend the methods used so far to
develop a better estimation method for the transport in these asymmetrical matrices.


% stotland Errata variable activation energy.. 
%%%%%%%%%
%\section{stotland's errata model}



%%%%%%%%%%%%%%%%%%%%  HEAT must appear!
%%%%%%%%%%%%%%%%%%%% I'm in charge of heat
%% bullets - where does this model appear - balls and springs (leads to heat) - capacitors
%%             - conductance hopping

\section{Heat transport}

\begin{comment}
Heat conduction by phonons is also affected by the localization properties
of the model. If disorder scatters normal modes and induces diffusive 
energy transport, followed by normal heat conduction, then in accordance to Fourier's law, 
the heat current $J$ depends inversely on the system size: $J\sim N^{-1}$. 
However, in recent studies \ref{who} it was shown that $J\sim N^{-\alpha}$ 
with $\alpha\ne 1$ is sometimes the case for disordered $1d$ harmonic chains.
We wish to investigate the validity of Fourier's law for quasi-$1d$ and 
higher dimensional disordered systems.
\end{comment}

As presented in the introduction, heat conduction by phonons has some
similarities with other transport properties. We wish to relate as much
as possible the various definitions of conductance coefficients, in 
particular the relation between the spectral analysis,
the resistor network computation based on Kirchoff's equation
and the covariance matrix formalism Lyapunov equation. Our initial focus
will be on banded sparse matrices, since these provide methods to distinguish
between percolative effects and disorder.


%%%%%%%%%
\section{The spectral properties of Sinai diffusion in $1d$}

In a symmetric $1d$ random rate system, the question of
diffusion is equivalent to the question of finite resistivity.
The total conductivity of serially connected conductors is their 
harmonic average, meaning that the diffusion is regular
if the harmonic average is bounded, i.e.\ if
%
\begin{align}
 \sum_n \frac{1}{w_n} < \infty 
\end{align}
%
On the other hand, for an asymmetric system, an activation energy 
builds up, and a Sinai diffusion behavior is observed.
The conductance is exponentially small in the length, and the spreading 
is anomalous. The difference between those two models is clearly highly 
significant. We wish to investigate the difference between the models,
and the manifestation of this difference in the spectral properties of the model.

%%%%%%%%%
\section{Sinai diffusion in $2d$}

Contrary to random walks in \emph{symmetric} random environments,
in which much is known in any dimension,
the general \emph{asymmetric} random walk in random environment is
largely an open question in more than one dimension. For example, 
the question of whether large enough fluctuations can allow
sub diffusion in high dimensions is still open.
Specifically we ask what is the change
in the spectrum upon introduction of asymmetry, and 
if this change allows sub diffusive modes.




%%%%%%%%%%
\section{Quantum spreading}


%%%%
Past studies in quantum spreading
\cite{cohen_wave_2000,izrailev_evolution_1997,venegas-andraca_quantum_2012,guarneri_spectral_1989,*guarneri_estimate_1993} concerned disordered systems, but
without taking strong \emph{sparsity} into account.
Our question is whether the resistor network picture,
which is vital in stochastic sparse networks, is also relevant to the quantum picture.


We shall begin with a simple model described by a real symmetric Hamiltonian:
%
\begin{align}
  \mathcal{H}_{nm} = \textrm{random}[\pm] \eexp{-\textrm{random}[x]} \qquad 0<|n-m|\le b
\end{align}
%
Where the parameter $x$ is distributed uniformly $x\in [0,\sigma]$. 

%%%%%%%%%% many body must appear after the quantum case

%%%%%%%%%
\section{The charge carrier discreteness}\label{sec:discreteness}

Our work so far considered probability rate equations and single-particle Hamiltonians, which are valid 
for a single particle in the network.
Having many non interacting particles is equivalent to having many 
realizations of a single particle, leaving the picture intact. 
However, if we have multiple \emph{interacting} particles in the network, 
additional factors may contribute. For example, a single occupancy rule 
(i.e.\ each site may have up to one particle) may cause Fermi blocking. 
%Another issue is that the single particle behavior might differ from
%the bulk behavior, e.g.\ it could be sub-diffusive 
%while the bulk spreading is diffusive \cite{richards_theory_1977,hung_diffusion_2012}.




% Kottos. Quantum version. Localization etc. 
